Well-quasi-orderings on word languages

### 2024-08-12 src/conclusion.tex
typos + amalgamation + conclusion

\subparagraph{Monoid equations}  It could be interesting to understand which
monoids $M$ recognize languages that are \kl{well-quasi-ordered} by the
\kl{infix}, \kl{prefix} or \kl{suffix} relations. This is connected to finding
which classes of graphs of \emph{bounded clique-width} are
\kl{well-quasi-ordered} with respect to the \emph{induced subgraph relation}, a
subject that is recently gaining traction.

>>>>

\subparagraph{Monoid equations}  It could be interesting to understand which
monoids $M$ recognize languages that are \kl{well-quasi-ordered} by the
\kl{infix}, \kl{prefix} or \kl{suffix} relations. This research direction is
connected to finding which classes of graphs of \emph{bounded clique-width} are
\kl{well-quasi-ordered} with respect to the \emph{induced subgraph relation},
as shown in \cite{DRT10}, and recently revisited by one of the authors in
\cite{L24:arxiv:v2}.

\subparagraph{Complexity} We have chosen to disregard complexity considerations
when proving decidability results, as we do not believe that a fined grained
complexity analysis would be particularly enlightening. However, we conjecture
that for regular languages, the complexity should be polynomial in the size of
the minimal automaton recognizing the language.

### 2024-08-12 src/infixes.tex
typos + amalgamation + conclusion

@@

\begin{corollary}
    \label{inf-period-union-chains:lem}
    Let $x,u,y \in \Sigma^*$ be words.  The following 
    are finite unions of \kl{chains} for the infix relation:
    $\InfPeriodChain{x}u$, $u \InfPeriodChain{x}$,
    and $\InfPeriodChain{x} u \InfPeriodChain{y}$.
\end{corollary}
\begin{proof}
    Because $\InfPeriodChain{x}$ is a finite union of \kl{chains} for the \kl{suffix}
    relation (using \cref{inf-period-chain:lem}), we conclude that $\InfPeriodChain{x}u$ is one too for
    the \kl{infix} relation. This holds similarly for $u \InfPeriodChain{x}$.
    In the case of $\InfPeriodChain{x} u \InfPeriodChain{y}$,
    we remark that $\InfPeriodChain{x} u$ is a finite union of chains
    for the \kl{suffix relation},
    and that $\InfPeriodChain{y}$ is one for the \kl{prefix relation}
    (using again \cref{inf-period-chain:lem}).
    As a consequence,
    $\InfPeriodChain{x}u \InfPeriodChain{y}$ is a finite union 
    of \kl{chains} for the \kl{infix} relation.
\end{proof}

@@

    Let us assume that $k \geq 2$ in the following proof for symmetry purposes,
    and argue later on that when $k = 1$ the same argument goes through.
    Because $L$ is \kl{well-quasi-ordered} by the \kl{infix relation}, there
    exists $i < j$ such that $w[\vec{n_i}]$ is an \kl{infix} of $w[\vec{n_j}]$.
    Now, because of the values of $\vec{n_j}$, there exists $1 \leq \ell \leq
    k-1$ such that $w[\vec{n_i}]$ is actually an \kl{infix} of $u_{\ell}
    v_{\ell}^{n_{j,\ell}} u_{\ell+1} v_{\ell+1}^{n_{j,\ell+1}} u_{\ell+2}$.
    Even more,
    one of the three following equations holds:
    \begin{itemize}
        \item $w[\vec{n_i}] \infleq v_{\ell}^{n_{j,\ell}} u_{\ell+1} v_{\ell+1}^{n_{j,\ell+1}}$,
        \item $w[\vec{n_i}] \infleq u_{\ell}
            v_{\ell}^{n_{j,\ell}}$,
        \item $w[\vec{n_i}] \infleq
            v_{\ell+1}^{n_{j,\ell+1}} u_{\ell+2}$.
    \end{itemize}
    In all those cases, we conclude using \cref{periodic-infixes:lem}.

@@

We are now ready to restate and proof our main theorem.

@@

\AP Let us now formally introduce the notion of \kl{amalgamation systems}, and
recall some results from \cite{ASZZ24} that will be useful for the proof of
\cref{infix-amalgamation:thm}. Let us recall that over an alphabet $(\Sigma,
=)$ a \kl{subword embedding} between two words $u \in \Sigma^*$ and $v \in
\Sigma^*$ is a function $\rho \colon \range{\card{u}} \to \range{\card{v}}$
such that $u_i = v_{\rho(i)}$ for all $i \in \range{\card{u}}$. We write
$\intro*\HigEmb(u,v)$ the set of all \kl{subword embeddings} between $u$ and
$v$. It may be useful to notice that the set of finite words over $\Sigma$
forms a category when we consider \kl{subword embeddings} as morphisms.

\AP Given a \kl{subword embedding} $f \colon u \to v$ between two words $u$ and
$v$, there exists a unique decomposition $v = \GapWord{f}{0} u_1 \GapWord{f}{1}
\cdots \GapWord{f}{k-1} u_k \GapWord{f}{k}$ where $\GapWord{f}{i} =
v[f(i)+1:f(i+1)-1]$ for all $1 \leq i \leq k-1$, $\GapWord{f}{k} =
v[f(k)+1:\card{v}]$, and $\GapWord{f}{0}   = v[1: f(1)-1]$.  We say that
$\intro*\GapWord{f}{i}$ is the $i$-th \intro{gap word} of $f$. Given a run $r
\in R$, and $i \in \range[0]{\card{\canrun(r)}}$, we define the \intro{gap
language} of $r$ at position $i$ $\GapLanguage{r}{i} \defined
\setof{\GapWord{f}{i}}{\exists s \in R. \exists f \in E(r,s) }$.

\begin{definition}
    An \intro{amalgamation system}
    is a triple $(\Sigma, R, E, \canrun)$ where
    $\Sigma$ is a finite alphabet,
    $R$ is a set of so-called \emph{runs},
    and 
    $E$ is a set of so-called \emph{run embeddings}.
    \begin{itemize}
        \item For all $r, s \in R$,
            $E(r,s) \subseteq \HigEmb(\canrun(r), \canrun(s))$.
        \item $(R, E)$ forms a category.
        \item $(R, \leq_E)$ is a well-quasi-ordered set.
        \item concatenative amalagamation
    \end{itemize}

@@

The notion of \kl{amalgamation system} is tailored to produce \emph{pumping
arguments}, which is exactly what our \cref{pumping-periods:lem} talks about.
Because the notion is quite abstract, it may be useful to replicate some
intuitions developed by the authors of \cite{ASZZ24}.

@@

    Let us first argue that languages $\GapLanguage{r}{i}$ are all \kl{chains}
    for the \kl{prefix ordering}. According to \cite[Lemma 4.1]{ASZZ24}, either
    $\GapLanguage{r}{i}$ is a \kl{chain} for the \kl{prefix relation}, or it is
    an \kl{unbounded language}. Note that \kl{unbounded languages}
    are \kl{antichains} for the \kl{infix relation}, and therefore
    \textbf{unclear}.

>>>>

\begin{corollary}
    \label{powers-infixes:cor}
    Let $u,v \in \Sigma^*$ and $k, \ell \in \Nat$
    such that $k \geq \factorial[p]{\card{u} \times \card{v}}$,
    $\ell \geq \factorial[p]{\card{v} \times \card{u}}$,
    and $u^k \infleq v^\ell$.
    Then, there exists $w \in \Sigma^*$ of size at most
    $\min \set{\card{u}, \card{v}}$ and a $p \in \Nat$
    such that
    $u^k \infleq v^\ell \infleq w^p$.
\end{corollary}

@@

\begin{corollary}
    \label{inf-period-union-chains:lem}
    Let $x,u,y \in \Sigma^*$ be words.  The following 
    are finite unions of \kl{chains} for the infix relation:
    $\InfPeriodChain{x}u$, $u \InfPeriodChain{x}$,
    and $\InfPeriodChain{x} u \InfPeriodChain{y}$.
\end{corollary}
\begin{proof}
    Because $\InfPeriodChain{x}$ is a finite union of \kl{chains} for the \kl{suffix}
    relation (rsing \cref{inf-period-chain:lem}), we conclude that $\InfPeriodChain{x}u$ is one too for
    the \kl{infix} relation. This holds similarly for $u \InfPeriodChain{x}$.
    In the case of $\InfPeriodChain{x} u \InfPeriodChain{y}$,
    we remark that $\InfPeriodChain{x} u$ is a finite union of chains
    for the \kl{suffix relation},
    and that $\InfPeriodChain{y}$ is one for the \kl{prefix relation}
    (using again \cref{inf-period-chain:lem}).
    As a consequence,
    $\InfPeriodChain{x}u \InfPeriodChain{y}$ is a finite union 
    of \kl{chains} for the \kl{infix} relation.
\end{proof}

@@

    Let us assume that $k \geq 2$ in the following proof for symmetry purposes,
    and argue later on that when $k = 1$ the same argument goes through.
    Because $L$ is \kl{well-quasi-ordered} by the \kl{infix relation}, there
    exists $i < j$ such that $w[\vec{n_i}]$ is an \kl{infix} of $w[\vec{n_j}]$.
    Now, because of the values of $\vec{n_j}$, there exists $1 \leq \ell \leq
    k-1$ such that $w[\vec{n_i}]$ is actually an \kl{infix} of $u_{\ell}
    v_{\ell}^{n_{j,\ell}} u_{\ell+1} v_{\ell+1}^{n_{j,\ell+1}} u_{\ell+2}$.
    Even more,
    one of the three following equations holds:
    \begin{itemize}
        \item $w[\vec{n_i}] \infleq v_{\ell}^{n_{j,\ell}} u_{\ell+1} v_{\ell+1}^{n_{j,\ell+1}}$,
        \item $w[\vec{n_i}] \infleq u_{\ell}
            v_{\ell}^{n_{j,\ell}}$,
        \item $w[\vec{n_i}] \infleq
            v_{\ell+1}^{n_{j,\ell+1}} u_{\ell+2}$.
    \end{itemize}
    In all those cases, we conclude using \cref{powers-infixes:cor}
    that there exists $x,y \in \Sigma^+$ of size at most $s$, and 
    a number $1 \leq t \leq k$ such that
    $v_i^{n_i} \in \InfPeriodChain{x}$ for all $1 \leq i \leq t$,
    and
    $v_i^{n_i} \in \InfPeriodChain{y}$ for all $t < i \leq k$.
    In particular,
    $w[\vec{n_i}] \in \InfPeriodChain{x} u_{t} \InfPeriodChain{y}$.

@@

We are now ready to restate and prove our main theorem.

@@

Let us now formally introduce the notion of \kl{amalgamation systems}, and
recall some results from \cite{ASZZ24} that will be useful for the proof of
\cref{infix-amalgamation:thm}. The notion of \kl{amalgamation system} is
tailored to produce \emph{pumping arguments}, which is exactly what our
\cref{pumping-periods:lem} talks about. At the core of a pumping argument,
there is a notion of \emph{run}, which could for instance be a sequence of
transitions taken in a finite state automaton. Continuing on the analogy with
finite automata, there is a natural ordering between runs, i.e., a run is
smaller than another one if one can ``unroll loops" of the first run to obtain
the second one. Typical pumping arguments then rely on the fact that
\emph{minimal} runs are of finite size, and that all other runs are
obtained by ``gluing" minimal runs together. This is exactly what
\kl{amalgamation systems} are about.

\AP Let us recall that over an alphabet $(\Sigma, =)$ a \kl{subword embedding}
between two words $u \in \Sigma^*$ and $v \in \Sigma^*$ is a function $\rho
\colon \range{\card{u}} \to \range{\card{v}}$ such that $u_i = v_{\rho(i)}$ for
all $i \in \range{\card{u}}$. We write $\intro*\HigEmb(u,v)$ the set of all
\kl{subword embeddings} between $u$ and $v$. It may be useful to notice that
the set of finite words over $\Sigma$ forms a category when we consider
\kl{subword embeddings} as morphisms. Given a \kl{subword embedding} $f \colon
u \to v$ between two words $u$ and $v$, there exists a unique decomposition $v
= \GapWord{f}{0} u_1 \GapWord{f}{1} \cdots \GapWord{f}{k-1} u_k \GapWord{f}{k}$
where $\GapWord{f}{i} = v[f(i)+1:f(i+1)-1]$ for all $1 \leq i \leq k-1$,
$\GapWord{f}{k} = v[f(k)+1:\card{v}]$, and $\GapWord{f}{0}   = v[1: f(1)-1]$.
We say that $\intro*\GapWord{f}{i}$ is the $i$-th \intro{gap word} of $f$.

\begin{definition}
    An \intro{amalgamation system}
    is a triple $(\Sigma, R, E, \canrun)$ where
    $\Sigma$ is a finite alphabet,
    $R$ is a set of so-called \emph{runs},
    and 
    $E$ is a set of so-called \emph{run embeddings},
    and $\canrun \colon R \to (\Sigma \cup \set{\varepsilon})^*$ is a 
    function computing a \emph{canonical decompostion} of a run.
    Given a run $r \in R$, and $i \in \range[0]{\card{\canrun(r)}}$, we define
    the \intro{gap language} of $r$ at position $i$ as $\GapLanguage{r}{i} \defined
    \setof{\GapWord{f}{i}}{\exists s \in R. \exists f \in E(r,s) }$.
    An \kl{amalgamation system} furthermore satisfies the following 
    properties:
    \begin{enumerate}
        \item For all $r, s \in R$,
            $E(r,s) \subseteq \HigEmb(\canrun(r), \canrun(s))$.
        \item $(R, E)$ forms a category.
        \item $(R, \leq_E)$ is a well-quasi-ordered set.
        \item For all $r, s, t \in R$,
            for all $f \in E(r,s)$,
            for all $g \in E(r,t)$,
            and for all $0 \leq i_0 \leq \card{\canrun(r)}$,
            there exists a run $z \in R$ and morphisms
            $h \in E(s,z)$ and $k \in E(t,z)$
            satisfying
            $h \circ f = k \circ g$,
            $\GapLanguage{h \circ f}{i} = \GapLanguage{f}{i} \GapLanguage{g}{i}$
            or 
            $\GapLanguage{h \circ f}{i} = \GapLanguage{g}{i} \GapLanguage{f}{i}$
            for every $0 \leq j \leq n$,
            and
            $\GapLanguage{h \circ f}{i_0} = \GapLanguage{f}{i_0} \GapLanguage{g}{i_0}$.
    \end{enumerate}

@@

Intuitively, the definition of an amalgamation system allows the comparison of
runs, and the proper ``gluing" of runs together to obtain new runs.

@@

    Let $u$ and $v$ be two words in the \kl{gap language}
    $\GapLanguage{r}{\ell}$ for some $r \in M$ and $0 \leq \ell \leq
    \card{\canrun(r)}$. We have corresponding runs $s_1, s_2 \in R$ and
    embeddings $f_1 \in E(r,s_1)$ and $f_2 \in E(r,s_2)$ such that
    $\GapWord{f_1}{\ell} = u$ and $\GapWord{f_2}{\ell} = v$. Using the
    amalgamation property, we can inductively construct runs $r_k \in R$ and
    embeddings $f_k \in E(r,s_k)$ such that $\GapWord{f_k}{\ell} = u v^k u$,
    and $\GapWord{f_k}{i}$ is obtained by concatenating in some order two
    occurrences of $\GapWord{f_1}{i}$ and $k$ occurrences of
    $\GapWord{f_2}{i}$. Because $L$ is \kl{well-quasi-ordered} for the
    \kl{infix relation}, there exists $i < j$ such that $\yieldrun(r_i) \infleq
    \yieldrun(r_j)$. Assuming that $\card{\canrun(r)}$ is greater than $2$, and
    that we have extracted an increasing sequence of indices so that
    $\alpha(i+1) > \factorial{\card{\yieldrun(r_i)}}$, we conclude that $u
    v^{\alpha(i)} u$ is an \kl{infix} of $\yieldrun(r_j)$.
    Because of the number of repetitions, a \emph{large} infix of $v^{\alpha(i)}$ must be
    an infix of some $\GapWord{f_2}{p}^{\alpha(j)}$.
    We conclude that $\Gapword{f_2}{p}^{\alpha(j)}$ and $v$
    have a common period. Refining the analysis, we also conclude
    that $uv^{q}$ and $\GapWord{f_2}{p}^{\alpha(j)}$ have the same period,
    which proves that $uv^{q}u$ belongs to some $\InfPeriodChain{x}$
    \textbf{todo}.

    \begin{quote}
        Idea: when pumping like this, we can always find
        large blocks of $\GapWord{f_2}{p}$, so this looks like
        something suitable for \cref{pumping-periods:lem}.
        This should allow us to conclude that every pair of words
        $u,v$ in a given gap language
        are infixes of \emph{the same} periodic word repeated.
    \end{quote}

### 2024-08-12 src/introduction.tex
typos + amalgamation + conclusion

We circumvent this issue by adding some regularity assumptions on the languages
$L \subseteq \Sigma^*$ that we consider. This can be done by assuming that $L$
is recognized by a finite automaton, or more generally that $L$ is the language
of an \kl{amalgamation system}. Under these relatively mild assumptions, we
also prove that \kl{well-quasi-ordered} languages under the \kl{infix relation}
are precisely the finite unions of \kl{chains}.

>>>>

We circumvent this issue by adding some regularity assumptions on the languages
$L \subseteq \Sigma^*$ that we consider. This can be done by assuming that $L$
is recognized by a finite automaton, or more generally that $L$ is the language
of an \kl{amalgamation system}, a recently introduced notion that abstracts the
notion of systems whose runs can be suitably ``pumped'' \cite{ASZZ24}. Under
this relatively mild assumption, we also prove that \kl{well-quasi-ordered}
languages under the \kl{infix relation} are precisely the finite unions of
\kl{chains}.
